---
author: "Nirmiti Dighe and Girish Kancharla"
title: "Space Objects in Earth's Orbit"
format: pdf
date: "`r Sys.Date()`"
editor: visual
---

#### Introduction -

The dataset we considered for the project is based on the space objects such as the satellites and debris present and launched into the Earth's orbit. The motivation behind choosing the dataset was since space junk is a real issue and yearly there are various satellites been released into the earth's orbits. There is wear and tear, collisions taking place which eventually turns the satellites into junks or debris. The space thus get accumulated and overpopulated with different objects over time which does need the necessary attention. We have visualized various questions being: The dataset described contains various attributes few are:

LAUNCH_DATE : Year in which the object was launched

APOAPSIS : Maximum distance from Earth

OBJECT_TYPE: If the object is payload, rocket body, debris or TBA(To be assigned)

RCS_SIZE: Radar cross section size category (Small, Medium, Large), indicating the size

The dataset analysis the space objects data from the year 1961 to 2021. We have visualized different plots such as the cumulative area plot, stacked bar graph, line plot and scatter plot.

### Visualization -

```{r}
# Loading the libraries
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r}
# Loading the dataset and cleaning

space_decay<-read.csv("space_decay.csv")
space_decay$DECAY_DATE<-NULL
colnames(space_decay)
space_decay<-na.omit(space_decay)
sum(space_decay == "", na.rm = TRUE)
unique(space_decay$LAUNCH_DATE)
unique(space_decay$COUNTRY_CODE)
```

#### 1. Where are the maximum number of space objects present, ofcourse right up in the space but.... where? Lower earth orbit, medium earth orbit or higher earth orbit, let us visualize it!

```{r}
space_decay$Orbit <- ifelse(space_decay$APOAPSIS <= 2000, "Low Earth Orbit",
                     ifelse(space_decay$APOAPSIS <= 36000, "Medium Earth Orbit", 
                            "High Earth Orbit"))

space_decay <- space_decay[space_decay$LAUNCH_DATE >= 1990 &
                             space_decay$LAUNCH_DATE <= 2021, ]
year_counts <- aggregate(rep(1, nrow(space_decay)),
                         by = list(Year = space_decay$LAUNCH_DATE, 
                                   Orbit = space_decay$Orbit),
                         FUN = sum)
colnames(year_counts)[3] <- "Count"
year_counts$Orbit <- factor(year_counts$Orbit, 
                            levels = c("High Earth Orbit", 
                                       "Medium Earth Orbit", "Low Earth Orbit"))

year_counts <- year_counts[order(year_counts$Orbit, year_counts$Year), ]
year_counts$Cumulative <- ave(year_counts$Count, year_counts$Orbit, 
                              FUN = cumsum)

ggplot(year_counts, aes(x = Year, y = Cumulative, fill = Orbit)) +
  geom_area(alpha = 0.8, size = 0.3) +
  labs(
    title = "Accumulation of Space Objects in Different Orbits (1990–2021)",
    x = "Year",
    y = "Cumulative Number of Objects",
    fill = "Orbit"
  ) +
  scale_fill_manual(values = c(
    "High Earth Orbit" = "cyan",
    "Medium Earth Orbit" = "brown",
    "Low Earth Orbit" = "darkkhaki"
  )) +
  scale_x_continuous(breaks = seq(1990, 2021, by = 5)) 

```

That looks like huge! With the data visualized for accumulation of space object in different orbits, we have seen it from the year 1990-2021 where we consider with the difference of 5 years each. Over the period of time the accumulation increases in the lower earth orbit with a good increase after 2010. The X-axis is scaled with the years followed by the cumulative number of objects on the y-axis. There is not much considerable increase in the amount of object accumulated in the medium earth orbit and high earth orbit which also comes to our notice that there are many space satellites be it for the telecommunication or technology as there are advancements in these industries we can see that the the number of objects launched has increased considerably.

#### 2. What kind of space objects are they? Are they all debris?

```{r}
#| label: object-type
#| fig.height: 5
#| fig.width: 5
#| fig.cap: "Barplot for different objects in space." 

ggplot(space_decay, aes(x = OBJECT_TYPE, fill = space_decay$OBJECT_TYPE)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none")
```

Visualizing it using a barchart displaying what kind of space objects are present in the space with how many in number. We can see that there are different types of space objects present such as the debris, payload, rocket body and to be addressed (TBA) objects. This barchart is enhanced into a stacked barchart to showcase the different sizes of space objects.

```{r}
#| label: object-type-size
#| fig.height: 5.2
#| fig.width: 5.3
#| fig.cap: "Stacked barplot of object types by size category (SMALL, MEDIUM, LARGE)."

valid_rows <- !is.na(space_decay$RCS_SIZE) &
              space_decay$RCS_SIZE %in% c("SMALL", "MEDIUM", "LARGE")
space_decay_subset <- space_decay[valid_rows, ]

stacked_data <- aggregate(
  x = list(Count = rep(1, nrow(space_decay_subset))),
  by = list(
    OBJECT_TYPE = space_decay_subset$OBJECT_TYPE,
    RCS_SIZE = space_decay_subset$RCS_SIZE
  ),
  FUN = sum
)

stacked_data$RCS_SIZE <- factor(stacked_data$RCS_SIZE, 
                      levels = c("SMALL", "MEDIUM", "LARGE"))

colors <- c(
  "SMALL" = "#C44E52",   
  "MEDIUM" = "#F2B134",  
  "LARGE" = "#86BBD8"  
)

ggplot(stacked_data, aes(x = OBJECT_TYPE, y = Count, fill = RCS_SIZE)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = colors) +
  labs(
    title = "Distribution of Space Objects by Type and Size Category",
    x = "Object Type",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 12),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.title = element_blank(),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  )
```

This displays how the debris, payload, rocketbody and TBA have different sizes objects been accumulated in the space. As we can see the debris which are small in size are huge in number whereas there is a portion of payload which are considered as payload. There are rocket bodies with medium and large objects with TBA being the least amount of small and medium objects being present.

#### 3. Let's see what percentages these different size objects are?

```{r}
space_decay_filtered <- space_decay %>%
  filter(!is.na(RCS_SIZE)) %>%
  filter(LAUNCH_DATE >= 1961, LAUNCH_DATE <= 2021) %>%
  filter(RCS_SIZE %in% c("SMALL", "MEDIUM", "LARGE"))

launch_by_size <- space_decay_filtered %>%
  group_by(RCS_SIZE) %>%
  summarise(Count = n()) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  mutate(RCS_SIZE = factor(RCS_SIZE, levels = c("SMALL", "MEDIUM", "LARGE")))  

ggplot(launch_by_size, aes(x = RCS_SIZE, y = Percent)) +
  geom_col(fill = "brown", width = 0.7) +
  geom_text(aes(label = paste0(round(Percent, 1), "%")), 
            hjust = -0.2, size = 5) +
  coord_flip() +
  labs(
    title = "Distribution of Space Objects by Size (1961–2021)",
    x = "Size Category",
    y = "Percentage"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    expand = expansion(mult = c(0, 0.2))
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )
```

The graph demonstrated showcases on the large, medium and small space objects with the percentage of these objects in total. We can see maximum percentage of small objects followed by large and medium.

#### 4. Different Countries that have launched the number of objects in space.

```{r}
space_decay_filtered <- space_decay %>%
  filter(!is.na(LAUNCH_DATE)) %>%
  filter(LAUNCH_DATE >= 2000 & LAUNCH_DATE <= 2021) %>%
  filter(!is.na(COUNTRY_CODE)) 

selected_countries <-   c("US", "PRC", "CIS", "IND")

space_decay_filtered <- space_decay_filtered %>%
  filter(COUNTRY_CODE %in% selected_countries)

launch_by_country_year <- space_decay_filtered %>%
  mutate(Launch_Year = as.numeric(LAUNCH_DATE)) %>%
  group_by(COUNTRY_CODE, Launch_Year) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(COUNTRY_CODE, Launch_Year) %>%
  group_by(COUNTRY_CODE) %>%
  mutate(Cumulative_Count = cumsum(Count))  # Cumulative launches over years

ggplot(launch_by_country_year, aes(x = Launch_Year, y = Cumulative_Count, 
                                   color = COUNTRY_CODE)) +
  geom_line(size = 1) +
  labs(
    title = "Country-wise space Objects Launched (2000–2021)",
    x = "Year",
    y = "Number of Objects",
    color = "Country"
  )
 
```

There were few countries that were considered to check the number of launches done over the period of time. The countries considered for the analysis were United States, China, Commonwealth of independent states, and India. We can see there is a considerable increase the launches over increasing years maximum for the United States followed by China leading it second, and then Russia and India respectively.

#### 5. What kind of relationship between perigee and apogee across different space objects?

```{r}
ggplot(space_decay, aes(x = PERIAPSIS, y = APOAPSIS, color = OBJECT_TYPE)) +
  geom_point(alpha = 0.6, size = 1.5) +
  labs(
    title = "Scatter Plot of Apogee vs Perigee by Object Type",
    x = "Perigee (km)",
    y = "Apogee (km)",
    color = "Object Type"
  )
```

It shows the relationship between perigee and apogee distances of space objects by type. Most of the debris and rocket bodies are present closer to earth, while payloads vary more. The spread indicates how different orbit types differ across different categories.

#### Conclusion -

Therefore, after visualizing different plots there is a clear understanding on the large number of space objects present currently orbiting Earth. The accumulation of the different space objects show there is a significant rise in space objects in lower earth orbit over time. The distribution by object type showcases that debris for the majority of the objects in space therefore space junk is considerably more and increasing. Such insights emphasize the concern around space congestion and the need for enhancement in debris management and satellite regulation.
